Associations between chlorophyll \emph{a} and various microcystin health advisory Concentrations

### 2016-06-01 manuscript/manuscript.tex
revisions entered into overleaf

\documentclass[11pt,]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
% Redefine \includegraphics so that, unless explicit options are
% given, the image width will not exceed the width of the page.
% Images get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\ScaleIfNeeded{%
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother
\let\Oldincludegraphics\includegraphics
{%
 \catcode`\@=11\relax%
 \gdef\includegraphics{\@ifnextchar[{\Oldincludegraphics}{\Oldincludegraphics[width=\ScaleIfNeeded]}}%
}%
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Associations between chlorophyll a and various microcystin health advisory Concentrations},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and advanced understanding required to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. Chlorophyll
\emph{a} is commonly measured, can be estimated \emph{in situ}, and has
been shown to be positively associated with microcystin. In this paper,
we use this association to provide estimates of chlorophyll \emph{a}
concentrations that are indicative of a higher probability of exceeding
select health advisory concentrations for microcystin. Using the 2007
National Lakes Assessment and a conditional probability approach, we
identify chlorophyll \emph{a} concentrations that are more likely than
not to be associated with an exceedance of a microcystin health advisory
level. We look at the recent US EPA health advisories for drinking water
as well as the World Health Organization levels for drinking water and
recreational use and identify a range of chlorophyll \emph{a}
thresholds. A 50\% chance of exceeding one of the specific advisory
microcystin concentrations of 0.3, 1, 1.6, and 2 \(\mu\)g/L is
associatied with chlorophyll \emph{a} concentration thresholds of 23,
64, 84, and 103, respectively. When managing for these various
microcystin levels, exceeding these reported chlorophyll \emph{a}
concentrations should be a trigger for further testing and possible
management action.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225
\(\mu\)g/L. Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent
the detection limits. Any value greater than that indicates the presence
of microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51\(\mu\)g/L and the mean was 3.17\(\mu\)g/L. Of all
lakes sampled, 21\% of lakes exceeded the USEPA Child level, 8.8\% of
lakes exceeded the USEPA Adult level, 11.7\% of lakes exceeded the WHO
Drinking level, and 7.3\% of lakes exceeded the WHO Recreational level.
Chlorophyll \emph{a}, ranged from 0.07 to 936 \(\mu\)g/L and this
captures the range of trrophic states from oligotrophic to
hypereutrophic. All lakes had detectable levels of chlorophyll \emph{a}.
The median concentration was 7.79 \(\mu\)g/L and the mean was 29.63
\(\mu\)g/L. The association between chlorophyll \emph{a} and the upper
confidence interval across a range of conditional probability values are
shown in Table \ref{tab:mc_chla_table}. Specific chlorophyll \emph{a}
that are associated with greater than even odds of exceeding the
advisory levels were 0.07, 0.07, 3, and 11\(\mu\)g/L for 0.3, 1.0, 1.6,
and 2.0 \(\mu\)g/L advisory levels, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

The chlorophyll \emph{a} cutoffs may be used to predict whether or not a
lake exceeds the microcystin health advisories. Doing so allows us to
compare the accuracy of the prediction as well as evaluate false
negatives. Total accuracy of the four cutoffs predicting microcystin
exceedances were 12\% for the USEPA children's drinking water advisory,
20\% for the WHO drinking water advisory, 22\% for the USEPA adult
drinking water advisory, and 3\% for the WHO regreational advisory
(Tables \ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table},
\ref{tab:adult_conmat_table}, \& \ref{tab:who_rec_conmat_table}).
However, total accuracy is only one part of the prediction performace
with which we are concerned.

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin exceedances, the error that should be avoided is predicting
that no exceedance has occurred when in fact it has. In other words, we
would like to avoid Type II errors and minimize the proportion of false
negatives. For the four chlorophyll \emph{a} cut-offs we had a
proportion of false negatives of 2\%, 2\%, 2\%, and 0\% for the USEPA
children's, the WHO drinking water, the USEPA adult, and the WHO
recreational advisories, respectively. In each case we missed less than
10\% of the lakes that in fact exceeded the microcystin advisory. While
this method performs well with regard to the false negative percentage,
it is possible that is a relic of the NLA dataset and testing with
additional data would allow us to confirm this result.

@@

Indeed, the probability of exceeding each of the four tested health
advisory levels increased as a function of chlorophyll \emph{a}
concentration (Figure \ref{fig:multi_cp_plot}). We used this association
to identify chlorophyll \emph{a} concentrations that were associated
with a range of probabilities of exceeding a given health advisory level
(Table \ref{tab:mc_chla_table}). For the purposes of this discussion we
focus on a conditional probability of 50\% or greater (i.e., greater
than even odds to exceed a health advisory level). The 50\% conditional
probability chlorophyll \emph{a} thresholds represents 59.1\%, 14.3\%,
5.2\%, and 93.9\% of sample lakes for the USEPA Child, the WHO Drinking,
the USEPA Adult, and the WHO recreational levels, respecitvely.

@@

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/chla_micro_scatter-1.pdf}
\caption{Scatterplot showing association betweeen chlorophyll \textit{a}
and microcystin-LR. \label{fig:chla_micro_scatter}}
\end{figure}

@@

\begin{verbatim}
## Loading required package: grid
\end{verbatim}

\begin{figure}[htbp]
\centering
\includegraphics{manuscript_files/figure-latex/epa_child_cp_plot-1.pdf}
\caption{Conditional probability plots showing association between the
probability of exceeding various microcystin health advisory Levels. A.)
Plot for USEPA Child (0.3 \(\mu\)g/L). B.) Plot for WHO Drinking (1
\(\mu\)g/L). C.) Plot for USEPA Adult (1.6 \(\mu\)g/L). D.) Plot for WHO
Recreational (2 \(\mu\)g/L). \label{fig:multi_cp_plot}}
\end{figure}

@@

\begin{longtable}[c]{@{}lllll@{}}
\caption{Chlorophyll \textit{a} concentrations that are associated with
a 50\% probability of exceeding a microcystin health advisory
concentration. \label{tab:mc_chla_table}}\tabularnewline
\toprule
Cond. Probability U & SEPA Child (0.3 \(\mu\)g/L) W & HO Drink (1
\(\mu\)g/L) U & SEPA Adult (1.6 \(\mu\)g/L) W & HO Recreational (2
\(\mu\)g/L)\tabularnewline
\midrule
\endfirsthead
\toprule
Cond. Probability U & SEPA Child (0.3 \(\mu\)g/L) W & HO Drink (1
\(\mu\)g/L) U & SEPA Adult (1.6 \(\mu\)g/L) W & HO Recreational (2
\(\mu\)g/L)\tabularnewline
\midrule
\endhead
0.1 & 0.07 & 0.07 & 0.07 & 1\tabularnewline
0.2 & 0.07 & 5 & 11 & 15\tabularnewline
0.3 & 3 & 18 & 32 & 39\tabularnewline
0.4 & 11 & 39 & 67 & 78\tabularnewline
0.5 & 23 & 64 & 84 & 103\tabularnewline
0.6 & 39 & 92 & 115 & 167\tabularnewline
0.7 & 65 & 116 & 274 & 274\tabularnewline
0.8 & 115 & 256 & 871 & 871\tabularnewline
0.9 & 138 & 318 & 871 & 871\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}llll@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the USEPA
childrens drinking water advisory.
\label{tab:child_conmat_table}}\tabularnewline
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endfirsthead
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endhead
Not Exceed & 344 & 78 & 422\tabularnewline
Exceed & 467 & 139 & 606\tabularnewline
Column Totals & 811 & 217 & 1028\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}llll@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the WHO
drinking water advisory.
\label{tab:who_drink_conmat_table}}\tabularnewline
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endfirsthead
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endhead
Not Exceed & 787 & 98 & 885\tabularnewline
Exceed & 120 & 23 & 143\tabularnewline
Column Totals & 907 & 121 & 1028\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}llll@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the USEPA adult
drinking water advisory. \label{tab:adult_conmat_table}}\tabularnewline
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endfirsthead
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endhead
Not Exceed & 897 & 82 & 979\tabularnewline
Exceed & 40 & 9 & 49\tabularnewline
Column Totals & 937 & 91 & 1028\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}llll@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the WHO
recreational water advisory.
\label{tab:who_rec_conmat_table}}\tabularnewline
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endfirsthead
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endhead
Not Exceed & 61 & 2 & 63\tabularnewline
Exceed & 892 & 73 & 965\tabularnewline
Column Totals & 953 & 75 & 1028\tabularnewline
\bottomrule
\end{longtable}

>>>>

\documentclass[11pt,]{article}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable,booktabs}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Associations between chlorophyll a and various microcystin health advisory Concentrations},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

Cyanobacteria harmful algal blooms (cHABs) are associated with a wide
range of adverse health effects that stem mostly from the presence of
cyanotoxins. To help protect against these impacts, several health
advisory levels have been set for some toxins. In particular, one of the
more common toxins, microcystin, has several advisory levels set for
drinking water and recreational use. However, compared to other water
quality measures, field measurements of microcystin are not commonly
available due to cost and advanced understanding required to interpret
results. Addressing these issues will take time and resources. Thus,
there is utility in finding indicators of microcystin that are already
widely available, can be estimated quickly and \emph{in situ}, and used
as a first defense against high levels of microcystin. Chlorophyll
\emph{a} is commonly measured, can be estimated \emph{in situ}, and has
been shown to be positively associated with microcystin. In this paper,
we use this association to provide estimates of chlorophyll \emph{a}
concentrations that are indicative of a higher probability of exceeding
select health advisory concentrations for microcystin. Using the 2007
National Lakes Assessment and a conditional probability approach, we
identify chlorophyll \emph{a} concentrations that are more likely than
not to be associated with an exceedance of a microcystin health advisory
level. We look at the recent US EPA health advisories for drinking water
as well as the World Health Organization levels for drinking water and
recreational use and identify a range of chlorophyll \emph{a}
thresholds. A 50\% chance of exceeding one of the specific advisory
microcystin concentrations of 0.3, 1, 1.6, and 2 \(\mu\)g/L is
associatied with chlorophyll \emph{a} concentration thresholds of 23,
68, 84, and 104 \(\mu\)g/L, respectively. When managing for these
various microcystin levels, exceeding these reported chlorophyll
\emph{a} concentrations should be a trigger for further testing and
possible management action.

@@

In the 2007 NLA, microcystin-LR concentrations ranged from 0.05 to 225
\(\mu\)g/L. Microcystin-LR concentrations of 0.05 \(\mu\)g/L represent
the detection limits. Any value greater than that indicates the presence
of microcystin-LR. Of those lakes with microcystin, the median
concentration was 0.51\(\mu\)g/L and the mean was 3.17\(\mu\)g/L. Of all
lakes sampled, 21\% of lakes exceeded the USEPA Child level, 8.8\% of
lakes exceeded the USEPA Adult level, 11.7\% of lakes exceeded the WHO
Drinking level, and 7.3\% of lakes exceeded the WHO Recreational level.
Chlorophyll \emph{a}, ranged from 0.07 to 936 \(\mu\)g/L and this
captures the range of trrophic states from oligotrophic to
hypereutrophic. All lakes had detectable levels of chlorophyll \emph{a}.
The median concentration was 7.79 \(\mu\)g/L and the mean was 29.63
\(\mu\)g/L. The association between chlorophyll \emph{a} and the upper
confidence interval across a range of conditional probability values are
shown in Table \ref{tab:mc_chla_table}. Specific chlorophyll \emph{a}
that are associated with greater than even odds of exceeding the
advisory levels were 23, 68, 84, and 104\(\mu\)g/L for 0.3, 1.0, 1.6,
and 2.0 \(\mu\)g/L advisory levels, respectively (Table
\ref{tab:mc_chla_table} \& Figure \ref{fig:multi_cp_plot}).

The chlorophyll \emph{a} cutoffs may be used to predict whether or not a
lake exceeds the microcystin health advisories. Doing so allows us to
compare the accuracy of the prediction as well as evaluate false
negatives. Total accuracy of the four cutoffs predicting microcystin
exceedances were 74\% for the USEPA children's drinking water advisory,
86\% for the WHO drinking water advisory, 89\% for the USEPA adult
drinking water advisory, and 91\% for the WHO regreational advisory
(Tables \ref{tab:child_conmat_table}, \ref{tab:who_drink_conmat_table},
\ref{tab:adult_conmat_table}, \& \ref{tab:who_rec_conmat_table}).
However, total accuracy is only one part of the prediction performace
with which we are concerned.

When using the chlorophyll \emph{a} cutoffs as an indicator of
microcystin exceedances, the error that should be avoided is predicting
that no exceedance has occurred when in fact it has. In other words, we
would like to avoid Type II errors and minimize the proportion of false
negatives. For the four chlorophyll \emph{a} cut-offs we had a
proportion of false negatives of 9\%, 8\%, 6\%, and 5\% for the USEPA
children's, the WHO drinking water, the USEPA adult, and the WHO
recreational advisories, respectively. In each case we missed less than
10\% of the lakes that in fact exceeded the microcystin advisory. While
this method performs well with regard to the false negative percentage,
it is possible that is a relic of the NLA dataset and testing with
additional data would allow us to confirm this result.

@@

Indeed, the probability of exceeding each of the four tested health
advisory levels increased as a function of chlorophyll \emph{a}
concentration (Figure \ref{fig:multi_cp_plot}). We used this association
to identify chlorophyll \emph{a} concentrations that were associated
with a range of probabilities of exceeding a given health advisory level
(Table \ref{tab:mc_chla_table}). For the purposes of this discussion we
focus on a conditional probability of 50\% or greater (i.e., greater
than even odds to exceed a health advisory level). The 50\% conditional
probability chlorophyll \emph{a} thresholds represents 28.6\%, 11\%,
8.9\%, and 7.2\% of sample lakes for the USEPA Child, the WHO Drinking,
the USEPA Adult, and the WHO recreational levels, respecitvely.

@@

\begin{verbatim}
## Saving 6 x 6.8 in image
\end{verbatim}

@@

@@

\begin{longtable}[c]{@{}lllll@{}}
\caption{Chlorophyll \textit{a} concentrations that are associated with
a 50\% probability of exceeding a microcystin health advisory
concentration. \label{tab:mc_chla_table}}\tabularnewline
\toprule
Cond. Probability & USEPA Child (0.3 \(\mu\)g/L) & WHO Drink (1
\(\mu\)g/L) & USEPA Adult (1.6 \(\mu\)g/L) & WHO Recreational (2
\(\mu\)g/L)\tabularnewline
\midrule
\endfirsthead
\toprule
Cond. Probability & USEPA Child (0.3 \(\mu\)g/L) & WHO Drink (1
\(\mu\)g/L) & USEPA Adult (1.6 \(\mu\)g/L) & WHO Recreational (2
\(\mu\)g/L)\tabularnewline
\midrule
\endhead
0.1 & 0.07 & 0.07 & 0.07 & 1\tabularnewline
0.2 & 0.07 & 4.00 & 12.00 & 17\tabularnewline
0.3 & 3.00 & 17.00 & 32.00 & 45\tabularnewline
0.4 & 11.00 & 37.00 & 68.00 & 77\tabularnewline
0.5 & 23.00 & 68.00 & 84.00 & 104\tabularnewline
0.6 & 39.00 & 97.00 & 115.00 & 185\tabularnewline
0.7 & 66.00 & 126.00 & 871.00 & 871\tabularnewline
0.8 & 116.00 & 271.00 & 871.00 & 871\tabularnewline
0.9 & 170.00 & 516.00 & 871.00 & 871\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}llll@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the USEPA
childrens drinking water advisory.
\label{tab:child_conmat_table}}\tabularnewline
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endfirsthead
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endhead
Not Exceed & 643 & 95 & 738\tabularnewline
Exceed & 168 & 122 & 290\tabularnewline
Column Totals & 811 & 217 & 1028\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}llll@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the WHO
drinking water advisory.
\label{tab:who_drink_conmat_table}}\tabularnewline
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endfirsthead
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endhead
Not Exceed & 841 & 78 & 919\tabularnewline
Exceed & 66 & 43 & 109\tabularnewline
Column Totals & 907 & 121 & 1028\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}llll@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the USEPA adult
drinking water advisory. \label{tab:adult_conmat_table}}\tabularnewline
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endfirsthead
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endhead
Not Exceed & 884 & 57 & 941\tabularnewline
Exceed & 53 & 34 & 87\tabularnewline
Column Totals & 937 & 91 & 1028\tabularnewline
\bottomrule
\end{longtable}

@@

\begin{longtable}[c]{@{}llll@{}}
\caption{Confusion matrix comparing chlorophyll \textit{a} predicted
exceedences (rows) versus real exceedances (columns) for the WHO
recreational water advisory.
\label{tab:who_rec_conmat_table}}\tabularnewline
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endfirsthead
\toprule
Not Exceed Exc & eed Row Tot & als &\tabularnewline
\midrule
\endhead
Not Exceed & 908 & 51 & 959\tabularnewline
Exceed & 45 & 24 & 69\tabularnewline
Column Totals & 953 & 75 & 1028\tabularnewline
\bottomrule
\end{longtable}
